\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2017-10-08 paper/casebase_jss.tex
add intro vignette and fix comp errors

\section{Case study 1: Veteran data (or ERSPC if we
can)}\label{case-study-1-veteran-data-or-erspc-if-we-can}

\begin{itemize}
\tightlist
\item
  First example
\item
  Show how we can test for non-proportional hazard?
\end{itemize}

\section{Case study 2: Bone-marrow
transplant}\label{case-study-2-bone-marrow-transplant}

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.82 & (0.46, 1.48)\\
\hline
Disease & 0.48 & (0.25, 0.92)\\
\hline
Phase (CR2 vs. CR1) & 1.43 & (0.56, 3.66)\\
\hline
Phase (CR3 vs. CR1) & 1.63 & (0.41, 6.45)\\
\hline
Phase (Relapse vs. CR1) & 4.73 & (2.13, 10.48)\\
\hline
Source & 1.51 & (0.49, 4.69)\\
\hline
Age & 0.99 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-6-1}

@@

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-6}
\end{figure}
\end{CodeChunk}

\section{Case study 3: Vaccination study (recurrent
events)}\label{case-study-3-vaccination-study-recurrent-events}

\begin{itemize}
\tightlist
\item
  Give a more complex example of sampling; time-dependent exposure

  \begin{itemize}
  \tightlist
  \item
    Sampling needs to be done manually, but fitting function can still
    be used
  \end{itemize}
\end{itemize}

\renewcommand\refname{Discussion}
\bibliography{casebase_references.bib}

>>>>

\section{Case study 1--European Randomized Study of Prostate Cancer
Screening
Data}\label{case-study-1european-randomized-study-of-prostate-cancer-screening-data}

This vignette introduces the main functions in the \texttt{casebase}
package. The methods implemented in this package are based on the method
developped in
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf}{Fitting
Smooth-in-Time Prognostic Risk Functions via Logistic Regression (Hanley
and Miettinen, 2009)}. A rigorous treatment of the theory is developed
in
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Saarela-2015-Lifetime_Data_Analysis.pdf}{A
case-base sampling method for estimating recurrent event intensities
(Saarela, 2015)} and
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Saarela_et_al-2015-Scandinavian_Journal_of_Statistics.pdf}{Non-parametric
Bayesian Hazard Regression for Chronic Disease Risk Assessment}. The
motivation for this work is nicely summarised by Cox:

\begin{figure}[htbp]
\centering
\includegraphics{figures/reid_cox.png}
\caption{}
\end{figure}

The purpose of the \texttt{casebase} package is to provide practitioners
with an easy-to-use software tool to predict the risk (or cumulative
incidence (CI)) of an event, for a particular patient. The following
points should be noted:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Time matching/risk set sampling (including Cox partial likelihood)
  eliminates the baseline hazard from the likelihood expression for the
  hazard ratios
\item
  If, however, the absolute risks are of interest, they have to be
  recovered using the semi-parametric Breslow estimator
\item
  Alternative approaches for fitting flexible hazard models for
  estimating absolute risks, not requiring this two-step approach? Yes!
  \href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf}{Hanley
  and Miettinen, 2009}
\end{enumerate}

\begin{quote}
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Hanley_Miettinen-2009-Inter_J_of_Biostats.pdf}{Hanley
and Miettinen, 2009} propose a fully parametric hazard model that can be
fit via logistic regression. From the fitted hazard function, cumulative
incidence and, thus, risk functions of time, treatment and profile can
be easily derived.
\end{quote}

The \texttt{casebase} package fits the family of hazard functions of the
form

\[ h(x,t) = exp[g(x,t)] \] where \(t\) denotes the numerical value
(number of units) of a point in prognostic/prospective time and \(x\) is
the realization of the vector \(X\) of variates based on the patient's
profile and intervention (if any). Different functions of \(t\) lead to
different parametric hazard models.

The simplest of these models is the one-parameter exponential
distribution which is obtained by taking the hazard function to be
constant over the range of \(t\).

\[ h(x,t) = exp(\beta_0 + \beta_1 x) \]

The instantaneous failure rate is independent of \(t\), so that the
conditional chance of failure in a time interval of specified length is
the same regardless of how long the individual has been on study a.k.a
the memoryless property (Kalbfleisch and Prentice, 2002).

The Gompertz hazard model is given by including a linear term for time:

\[ h(x,t)  = exp(\beta_0 + \beta_1 t + \beta_2 x) \]

Use of \(log(t)\) yields the Weibull hazard which allows for a power
dependence of the hazard on time (Kalbfleisch and Prentice, 2002):

\[ h(x, t)  = exp(\beta_0 + \beta_1 \log(t) + \beta_2 x) \]

Recall that the relative risk model (Cox, 1972)

\[ \lambda(t;x) = \lambda_0(t) exp(\mathbf{X}\boldsymbol{\beta})  \]
where \(\lambda_0(\cdot)\) is an arbitrary unspecified baseline hazard
function for continous time.

In the following table we provide a comparison between the Cox model and
case-base sampling:

\href{https://www.fields.utoronto.ca/programs/scientific/14-15/biomarker/slides/saarela.pdf}{Slides
from Olli Saarela}

We fist load the required packages:

Throughout this vignette, we make use of the European Randomized Study
of Prostate Cancer Screening data which ships with the \texttt{casebase}
package:

\begin{CodeInput}
R> data(ERSPC)
R> head(ERSPC)
\end{CodeInput}

\begin{CodeOutput}
  ScrArm Follow.Up.Time DeadOfPrCa
1      1         0.0027          0
2      1         0.0027          0
3      1         0.0027          0
4      0         0.0027          0
5      0         0.0027          0
6      0         0.0027          0
\end{CodeOutput}

\begin{CodeInput}
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R+                        levels = c(0,1), 
R+                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

The results of this study were published by
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Schroder_et_al-2009-NEJM.pdf}{Schroder
FH, et al. N Engl J Med 2009}. There's a really interesting story on how
this data was obtained. See \texttt{help(ERSPC)} and
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Liu_et_al-2015-Systematic_Reviews.pdf}{Liu
Z, Rich B, Hanley JA, Recovering the raw data behind a non-parametric
survival curve. Systematic Reviews 2014} for further details.

Population time plots can be extremely informative graphical displays of
survival data. They should be the first step in your exploratory data
analyses. We facilitate this task in the \texttt{casebase} package using
the \texttt{popTime} function. We first create the necessary dataset for
producing the population time plots:

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}
\end{CodeChunk}

We can see its contents and its class:

\begin{CodeInput}
R> head(pt_object)
\end{CodeInput}

\begin{CodeOutput}
            ScrArm   time event original.time original.event event status
1: Screening group 0.0027     0        0.0027              0     censored
2: Screening group 0.0027     0        0.0027              0     censored
3: Screening group 0.0027     0        0.0027              0     censored
4:   Control group 0.0027     0        0.0027              0     censored
5:   Control group 0.0027     0        0.0027              0     censored
6:   Control group 0.0027     0        0.0027              0     censored
   ycoord yc n_available
1: 159893  0           0
2: 159892  0           0
3: 159891  0           0
4: 159890  0           0
5: 159889  0           0
6: 159888  0           0
\end{CodeOutput}

\begin{CodeInput}
R> class(pt_object)
\end{CodeInput}

\begin{CodeOutput}
[1] "popTime"    "data.table" "data.frame"
\end{CodeOutput}
\end{CodeChunk}

The \texttt{casebase} package has a \texttt{plot} method for objects of
class \texttt{popTime}:

\begin{CodeInput}
R> plot(pt_object)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-5-1} \end{center}

We can also create exposure stratified plots by specifying the
\texttt{exposure} argument in the \texttt{popTime} function:

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R+                                      event = "DeadOfPrCa", 
R+                                      exposure = "ScrArm")
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}
\end{CodeChunk}

We can see its contents and its class:

\begin{CodeInput}
R> head(pt_object_strat)
\end{CodeInput}

\begin{CodeOutput}
$data
                 ScrArm    time event original.time original.event
     1:   Control group  0.0027     0        0.0027              0
     2:   Control group  0.0027     0        0.0027              0
     3:   Control group  0.0027     0        0.0027              0
     4:   Control group  0.0027     0        0.0027              0
     5:   Control group  0.0137     0        0.0137              0
    ---                                                           
159889: Screening group 14.9405     0       14.9405              0
159890: Screening group 14.9405     0       14.9405              0
159891: Screening group 14.9405     0       14.9405              0
159892: Screening group 14.9405     0       14.9405              0
159893: Screening group 14.9405     0       14.9405              0
        event status ycoord yc n_available
     1:     censored  88232  0           0
     2:     censored  88231  0           0
     3:     censored  88230  0           0
     4:     censored  88229  0           0
     5:     censored  88228  0           0
    ---                                   
159889:     censored      5  0           0
159890:     censored      4  0           0
159891:     censored      3  0           0
159892:     censored      2  0           0
159893:     censored      1  0           0

$exposure
[1] "ScrArm"
\end{CodeOutput}

\begin{CodeInput}
R> class(pt_object_strat)
\end{CodeInput}

\begin{CodeOutput}
[1] "popTimeExposure" "list"           
\end{CodeOutput}
\end{CodeChunk}

The \texttt{casebase} package also has a \texttt{plot} method for
objects of class \texttt{popTimeExposure}:

\begin{CodeInput}
R> plot(pt_object_strat)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-8-1} \end{center}

We can also plot them side-by-side using the \texttt{ncol} argument:

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-9-1} \end{center}

We first fit a Cox model, examine the hazard ratio for the screening
group (relative to the control group), and plot the cumulative incidence
function (CIF).

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R+                              data = ERSPC)
R> (sum_cox_model <- summary(cox_model))
\end{CodeInput}

\begin{CodeOutput}
Call:
survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
    ScrArm, data = ERSPC)

  n= 159893, number of events= 540

                        coef exp(coef) se(coef)     z Pr(>|z|)  
ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

                      exp(coef) exp(-coef) lower .95 upper .95
ScrArmScreening group     0.801       1.25     0.674     0.952

Concordance= 0.519  (se = 0.011 )
Rsquare= 0   (max possible= 0.075 )
Likelihood ratio test= 6.45  on 1 df,   p=0.0111
Wald test            = 6.37  on 1 df,   p=0.0116
Score (logrank) test = 6.39  on 1 df,   p=0.0115
\end{CodeOutput}
\end{CodeChunk}

We can plot the CIF for each group:

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-11-1} \end{center}

We compare it to the figure in
\href{https://github.com/sahirbhatnagar/casebase/blob/master/references/Schroder_et_al-2009-NEJM.pdf}{Schroder
FH, et al. N Engl J Med 2009} and see that the plots are very similar,
as is the hazard ratio and 95\% confidence interval:

\begin{figure}[htbp]
\centering
\includegraphics{figures/schroderNEJM_figure2.png}
\caption{}
\end{figure}

Next we fit several models using case-base sampling. The models we fit
differ in how we choose to model time.

The \texttt{fitSmoothHazard} function provides an estimate of the hazard
function \(h(x, t)\) is the hazard function, \(t\)) denotes the
numerical value (number of units) of a point in prognostic/prospective
time and \(x\) is the realization of the vector \(X\) of variates based
on the patient's profile and intervention (if any).

\begin{CodeInput}
R> # set the seed for reproducible output
R> set.seed(1234)
R> 
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R+                                                   data = ERSPC, 
R+                                                   ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-0.458  -0.458  -0.409  -0.409   2.246

Coefficients:
                      Estimate Std. Error z value            Pr(>|z|)    
(Intercept)            -7.7553     0.0584 -132.87 <0.0000000000000002 ***
ScrArmScreening group  -0.2382     0.0921   -2.59              0.0097 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3612.3  on 5938  degrees of freedom
AIC: 3616

Number of Fisher Scoring iterations: 5
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
ScrArmScreening group 
                 0.79 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
 2.5 % 97.5 % 
  0.66   0.94 
\end{CodeOutput}
\end{CodeChunk}

The \texttt{absoluteRisk} function provides an estimate of the
cumulative incidence curves for a specific risk profile using the
following equation:

\[ CI(x, t) = 1 - exp\left( - \int_0^t h(x, u) \textrm{d}u \right) \]

In the plot below, we overlay the estimated CIF from the casebase
exponential model on the Cox model CIF:

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-13-1} \end{center}

As we can see, the exponential model is not a good fit. Based on what we
observed in the population time plot, where more events are observed
later on in time, this poor fit is expected. A constant hazard model
would overestimate the cumulative incidence earlier on in time, and
underestimate it later on, which is what we see in the cumulative
incidence plot. This example demonstrates the benefits of population
time plots as an exploratory analysis tool.

Next we enter time linearly into the model:

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R+                                  data = ERSPC, 
R+                                  ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_time)
\end{CodeInput}

\begin{CodeOutput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-1.043  -0.487  -0.365  -0.275   2.744

Coefficients:
                      Estimate Std. Error z value            Pr(>|z|)    
(Intercept)            -9.0421     0.1183  -76.45 <0.0000000000000002 ***
Follow.Up.Time          0.2254     0.0158   14.24 <0.0000000000000002 ***
ScrArmScreening group  -0.2611     0.0940   -2.78              0.0055 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3394.9  on 5937  degrees of freedom
AIC: 3401

Number of Fisher Scoring iterations: 5
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
          (Intercept)        Follow.Up.Time ScrArmScreening group 
              0.00012               1.25285               0.77023 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
                         2.5 %  97.5 %
(Intercept)           0.000093 0.00015
Follow.Up.Time        1.214840 1.29263
ScrArmScreening group 0.639934 0.92534
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-15-1} \end{center}

We see that the Weibull model leads to a better fit.

Next we try to enter a smooth function of time into the model using the
\texttt{splines} package

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R+                                     data = ERSPC, 
R+                                     ratio = 100)
\end{CodeInput}

\begin{CodeOutput}
'Follow.Up.Time' will be used as the time variable
\end{CodeOutput}

\begin{CodeInput}
R> summary(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}

Call:
glm(formula = formula, family = binomial, data = sampleData)

Deviance Residuals: 
   Min      1Q  Median      3Q     Max  
-1.197  -0.522  -0.398  -0.212   3.187

Coefficients:
                      Estimate Std. Error z value             Pr(>|z|)    
(Intercept)           -10.5045     0.3342  -31.43 < 0.0000000000000002 ***
bs(Follow.Up.Time)1     4.8348     0.8400    5.76     0.00000000862087 ***
bs(Follow.Up.Time)2     1.8367     0.4859    3.78              0.00016 ***
bs(Follow.Up.Time)3     5.0478     0.6998    7.21     0.00000000000055 ***
ScrArmScreening group  -0.1503     0.0941   -1.60              0.11013    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for binomial family taken to be 1)

    Null deviance: 3619.1  on 5939  degrees of freedom
Residual deviance: 3356.5  on 5935  degrees of freedom
AIC: 3367

Number of Fisher Scoring iterations: 6
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
          (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
             0.000027            125.811140              6.275672 
  bs(Follow.Up.Time)3 ScrArmScreening group 
           155.673063              0.860457 
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
Waiting for profiling to be done...
\end{CodeOutput}

\begin{CodeOutput}
                          2.5 %     97.5 %
(Intercept)            0.000014   0.000051
bs(Follow.Up.Time)1   25.404219 687.891148
bs(Follow.Up.Time)2    2.425559  16.399777
bs(Follow.Up.Time)3   39.667300 626.079069
ScrArmScreening group  0.714859   1.033843
\end{CodeOutput}
\end{CodeChunk}

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = F, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]))
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-17-1} \end{center}

It looks like the best fit.

Since we are in the GLM framework, we can easily test for which model
better fits the data using a Likelihood ratio test (LRT). The null
hypothesis here is that the linear model is just as good as the larger
(in terms of number of parameters) splines model.

\begin{CodeInput}
R> anova(casebase_time, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
Analysis of Deviance Table

Model 1: DeadOfPrCa ~ Follow.Up.Time + ScrArm + offset(offset)
Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
  Resid. Df Resid. Dev Df Deviance     Pr(>Chi)    
1      5937       3395                             
2      5935       3357  2     38.4 0.0000000046 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

We see that splines model is the better fit.

\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}

@@

\begin{tabular}{l|r|l}
\hline
Covariates & HR & 95\% CI\\
\hline
Sex & 0.78 & (0.43, 1.41)\\
\hline
Disease & 0.41 & (0.22, 0.77)\\
\hline
Phase (CR2 vs. CR1) & 1.27 & (0.5, 3.23)\\
\hline
Phase (CR3 vs. CR1) & 1.77 & (0.44, 7.09)\\
\hline
Phase (Relapse vs. CR1) & 5.17 & (2.34, 11.43)\\
\hline
Source & 1.87 & (0.59, 5.88)\\
\hline
Age & 1.00 & (0.97, 1.02)\\
\hline
\end{tabular}

@@

{\centering \includegraphics{casebase_jss_files/figure-latex/unnamed-chunk-24-1}

@@

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:unnamed-chunk-24}
\end{figure}
\end{CodeChunk}

\section{Case study 3}\label{case-study-3}

\begin{itemize}
\tightlist
\item
  Glmnet, gam, and gbm
\end{itemize}

\section{Discussion}\label{discussion}

\begin{itemize}
\tightlist
\item
  Discuss
\end{itemize}

\bibliography{casebase_references.bib}
